Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-08-07 writeup/conStruct.tex
minor edits

\gc{
\paragraph{Relationship of this model to non-spatial structure models}

A nice feature of our approach is that our model contains a non-spatial assignment model as a special case. 
Setting $\alpha_0^{(k)}$ to zero for all $k$, we obtain a non-spatial model where
each cluster has its own allele frequency at each SNP, and individuals
draw a proportion of their ancestry from each cluster. This model is
strongly conceptually similar to that of STRUCTURE (and related
approaches ADMIXTURE), the only difference being that our likelihood
assumes that allele frequencies are normally distributed around their
expectations, while the standard assignment methods assume that the
error is binomial distributed (see XXXX). Therefore, to compare the fit
of

} 
\gb{support for the spatial model over the nonspatial model at a given value of $K$ 
indicates that isolation-by-distance is likely a feature of the data at that level.}

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, why
there is no true (or right) number of clusters for real datasets. 
%There are a number of reasons, both statistical and biological, 
%why choosing a single, ``best" number of clusters to describe the data may not be desirable.
%, including ambiguity in what exactly it means for any given value of $K$ to be ``best."
We discuss these reasons further in the Discussion below. In practice
it is useful to explore the implications of clustering at a variety of
different levels (i.e. across multiple $K$) for biological
intepretation and downstream analyses. 
However, we also recognize that in many circumstances it is 
useful to know the number of clusters with highest statistical
support in the (current) dataset.  %, for example to focus su 
%However, we also recognize that in many circumstances 
i%t is either useful or convenient 
%to present only a single set of results from a clustering analysis with a particular number of clusters.
%ideally a number large enough to describe meaningful partitions in the data, 
%but not so large as to be difficult to parse visually.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$.

Several approaches have been used as model choice criteria for the number of
clusters in \gc{population genetic} data, including: 
the use of information theoretic criteria \citep[e.g.][]{ADMIXTURE}, such as AIC or BIC;
comparisons of the likelihood of the data across different values of $K$, 
with various criteria on how to choose a single value (e.g., \citep{Evanno2005});
or comparisons of the marginal likelihood, 
generated either via various approximations \citep[e.g.][]{STRUCTURE}
or via a more rigorous and computationally intensive approach such as thermodynamic integration \citep{verity_nichols2016}
or inference using a Dirichlet process prior \citep{huelsenbeck2007inference}
See \cite{verity_nichols2016} for a discussion of these approaches and comparison
between several methods. %A comprehensive review of these approaches and their various merits is beyond the scope of this paper,
%but

\gc{IS PREDICTION accuracy the right phrase?}
We choose to take a cross-validation approach.  We estimate the
posterior on the parameters of the model using a ``training" partition
of the data, in practice the training set is a random subset of $90\%$
of the loci. We then calculate the likelihood of the
``testing" subset, the remaining $10\%$ of the loci, averaged over the
posterior of the parameters. We repeat this procedure a number of
times, take an average of the relative log-likelihood over the runs at
sumarize each $K$. 
and selecting the model with the significantly best predictive accuracy
(see Section \ref{Xvalidation} for more on our cross-validation procedure).
This approach rewards models with larger values of $K$, which increase
the likelihood of the training partition, as they fit the data better,
but the improvement in the likelihood will plateau (or peak) as above
a certain $K$ we are fitting noise specific to the training data and
so not fitting the testing set any better, 
%up to the point that cluster membership is invoked to model noise specific to the training data, 
%at which point predictive accuracy decreases, 
%and models with additional clusters are penalized.

@@

Trees in the genus \textit{Populus} (poplars, aspens, and cottonwoods) 
are distributed throughout the Northern Hemisphere, 
and include species of ecological and economic importance.
Species in the genus regularly co-occur and, 
where they do, they frequently hybridize (ref).
In addition, many \textit{Populus} species display patterns of isolation by distance, 
making the system well-suited to an analysis designed to tease apart 
genetic structure partitioned discretely across population (or species) clusters,
and continuously with geographic distance.

Here, we focus on two \textit{Populus} species in the Pacific Northwest: 
\textit{Populus trichocarpa}, the black cottonwood,
and \textit{Populus trichocarpa}, the balsam poplar.
These species have a broad zone of overlap and are known to hybridize \citep{geraldes_etal_2014}.

We address this question using a dataset generated and originally published by \cite{geraldes_etal_2014}.
The published data consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera}).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data using both the spatial and nonspatial \texttt{conStruct} models, 
running on $K = 1$ through 7,
and compared these models using the cross-validation procedure described above. 
The results of these analyses are shown in 
Figs \ref{populus_xvals}-\ref{populus_nsp_clst_covs}. %,populus_nsp_pies,populus_sp_clst_covs,populus_nsp_clst_covs

@@

We thank Marjorie Weber, William Wetzel, Mariah Meek, Yaniv Brandvain, and Matthew Stephens 
for invaluable comments on the method and manuscript, 
as well as Quentin Cronk, who helped interpret the \textit{Populus} analyses, 
and the attendees at the 2017 SSE Meeting in Portland, OR, whose votes determined the name of the method.
This work was supported in part by 
the National Science Foundation under award number NSF \#1262645 (DBI) to PR and GC, 
the National Institute of General Medical Sciences of the National Institutes of Health under award numbers NIH RO1GM83098 and RO1GM107374 to GC,
and the National Science Foundation under award numbers NSF \# 1148897 and \# 1402725 to GB.

@@

\section{Models, parameters, and priors}
\paragraph{Spatial vs. nonspatial}
In this paper, we discuss two types of models, spatial and nonspatial.
The spatial model is parameterized as in Eqn \ref{admixed_spatial_cov},
and the nonspatial model is described in Eqn \ref{admixed_discrete_covariance}.
The nonspatial model therefore has $3K$ fewer parameters than the spatial model,
as there are three $\alpha$ parameters that describe the continuous differentiation effect of distance per cluster,
and they are not included in the nonspatial model.

\paragraph{Single cluster}
Each of these models can be run with a single cluster ($K=1$), 
which involves a slight re-parameterization, detailed below.

>>>>

\paragraph{Relationship of this model to non-spatial structure models}

A nice feature of our approach is that the model described in Eqn \refeq{cross_cluster_covariance} 
contains a non-spatial assignment model as a special case 
(see \ref{model_app} for a more in-depth discussion). 
Setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a non-spatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomial distributed (see XXXX).  
Therefore, we can evaluate the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework.

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, 
why there is no true (or right) number of clusters for real datasets. 
We discuss these reasons further in the Discussion below. 
In practice, it is useful to explore the implications of clustering at a variety of different levels (
i.e. across multiple $K$) for biological interpretation and downstream analyses. 
However, we also recognize that in many circumstances it is 
useful to know the number of clusters with highest statistical
support in the (current) dataset.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$ 
and between the spatial and non-spatial models.

Several approaches have been used as model choice criteria 
for the number of clusters in population genetic data, including: 
the use of information theoretic criteria \citep[e.g.][]{ADMIXTURE}, such as AIC or BIC;
comparisons of the likelihood of the data across different values of $K$, 
with various criteria on how to choose a single value (e.g., \citep{Evanno2005});
or comparisons of the marginal likelihood, 
generated either via various approximations \citep[e.g.][]{STRUCTURE}
or via a more rigorous and computationally intensive approach such as thermodynamic integration \citep{verity_nichols2016}
or inference using a Dirichlet process prior \citep{huelsenbeck2007inference}
See \cite{verity_nichols2016} for a discussion of these approaches and comparison
between several methods.

\gc{IS PREDICTION accuracy the right phrase?}
We choose to take a cross-validation approach.
We estimate the posterior on the parameters of the model using a ``training" partition of the data 
(in practice the training set is a random subset of $90\%$ of the loci). 
We then calculate the likelihood of the ``testing" subset (the remaining $10\%$ of the loci), 
averaged over the posterior of the parameters. 
We repeat this procedure a number of times, 
take an average of the relative log-likelihoods over the runs at each $K$, 
and select the model with the significantly best predictive accuracy
(see Section \ref{Xvalidation} for more on our cross-validation procedure).
This approach rewards models with larger values of $K$, 
which increase the likelihood of the training partition, 
as they fit the data better,
but the improvement in the likelihood will plateau (or peak),
 as above a certain $K$ the model is fitting noise specific to the training data 
 and therefore not fitting the testing set any better.

@@

Trees in the genus \textit{Populus} (poplars, aspens, and cottonwoods) 
are distributed throughout the Northern Hemisphere, 
and include species of ecological and economic importance.
Species in the genus regularly co-occur and, 
where they do, they frequently hybridize (ref).

Here, we focus on two \textit{Populus} species in the Pacific Northwest: 
\textit{Populus trichocarpa}, the black cottonwood,
and \textit{Populus trichocarpa}, the balsam poplar.
These species have a broad zone of overlap and are hypothesized to hybridize \citep{geraldes_etal_2014}.
Both species are sampled over a large geographic region, 
making the system well-suited to an analysis designed to tease apart 
proportion of genetic structure that is partitioned discretely across population (or species) clusters,
as well as continuously with geographic distance.

We address this question using a dataset generated and originally published by \cite{geraldes_etal_2014}.
The published data consist of 434 individuals sampled from 35 drainages 
and genotyped at just over 33,000 loci (map of the sampling shown in Fig \ref{populus_map}).
The data were generated using RAD-seq, 
and showed a strong pattern of bias in allelic dropout 
(the majority of missing data were from drainages with only \textit{Populus balsamifera} individuals).
To ameliorate some of the problems that arise when there is a strong bias in which data are missing, 
we dropped loci for which any data were missing, 
resulting in just over 20,200 loci retained for analysis.  
We then analyzed these data using both the spatial and nonspatial \texttt{conStruct} models, 
running on $K = 1$ through 7,
and compared these models using the cross-validation procedure described above. 
The results of these analyses are shown in 
Figs \ref{populus_xvals}-\ref{populus_nsp_clst_covs}. %,populus_nsp_pies,populus_sp_clst_covs,populus_nsp_clst_covs

@@

We thank Marjorie Weber, William Wetzel, Mariah Meek, Yaniv Brandvain, and Matthew Stephens 
for invaluable comments on the method and manuscript, 
as well as Quentin Cronk, who provided input on the \textit{Populus} analyses, 
and the attendees at the 2017 SSE Meeting in Portland, OR, 
whose votes determined the name of the method.
This work was supported in part by 
the National Science Foundation under award number NSF \#1262645 (DBI) to PR and GC, 
the National Institute of General Medical Sciences of the National Institutes of Health under award numbers NIH RO1GM83098 and RO1GM107374 to GC,
and the National Science Foundation under award numbers NSF \# 1148897 and \# 1402725 to GB.

@@

\section{Models, parameters, and priors}\label{model_app}
\paragraph{Spatial vs. nonspatial}
In this paper, we discuss two types of models, spatial and nonspatial, 
each of which can be implemented with different numbers of clusters.
The spatial model is parameterized as in Eqn \ref{admixed_spatial_cov},
and the nonspatial model is described in Eqn \ref{admixed_discrete_covariance}.
The nonspatial model is a special case of the spatial model, 
in which the value of all $\alpha$ parameters is set to 0.
The nonspatial model therefore has $3K$ fewer parameters than the spatial model,
as there are three $\alpha$ parameters that describe the continuous differentiation effect of distance per cluster,
and they are fixed in the nonspatial model.

\paragraph{Single cluster}
Each of these models can be run with a single cluster ($K=1$), 
in which case the cluster-specific covariance parameter $\mu^{(k})$ 
and the global covariance parameter $\gamma$ become redundant.
The single-cluster model is therefore a special case of the multi-cluster model, 
in which we set $\mu$ to zero, as described below:
